DCMIP2016, Part 1: Models and Equation Sets

### 2017-02-18 DCMIP2016-Part1-Models.tex
Cleanup and updates

\Author[18]{Elijah}{Goodfriend}
\Author[18]{Hans}{Johansen}

@@

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical. DYNAMICO's design uniquely combines:
 \begin{itemize}
\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry,\citep{Dubos2015DYNAMICO10}
\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using builing blocks from the litterature,\citep{Thuburn2009Numerical,Ringler2010Unified}, and yielding, to a certain extent, an energy-conserving spatial discretization, and
\item a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation). 
\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

The GEM model, which is used for operational forecasting at Environment Canada, uses the Yin-Yang grid and horizontal discretization on an Arakawa C grid.  The vertical coordinate is of a log-hydrostatic-pressure type and vertical discretization is based on the Charney-Phillips grid.  A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.  A scalar viscosity is employed for wind components and tracers 
via an application of (\ref{eq:scalar_viscosity_direct}).  Viscosity operations are applied after the completion of the dynamic time step. {\color{red}[Perhaps expand this text a little bit?]}

@@

{\color{red}[SKAMAROCK]}

@@

The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical staggered nodal finite element method based on the cubed-sphere grid with terrain-following height-based coordinate.  The standard Eulerian equations are employed with moist density $\rho$, thermodynamic closure $\theta_v$ and tracer density $\rho q$.  These continuous equations are given in section \ref{sec:TempestEquations}.  The implementation includes both fully explicit time integration, using the UKG53 scheme described in section \ref{sec:UKG53Scheme}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.  Scalar hyperviscosity is employed for $\rho$, $\theta$ and tracer variables via repeated application of (\ref{eq:scalar_viscosity_direct}).  Vector hyperviscosity is also applied by decomposing the horizontal vector Laplacian into divergence damping (\ref{eq:divergence_damping}) and vorticity damping (\ref{eq:vorticity_damping}) terms.  Both viscosity operations are applied after the completion of all Runge-Kutta sub-cycles.

@@

\section{Moist Non-hydrostatic Equation Sets} \label{sec:EquationSets}

In this section we describe the fluid equations utilized by non-hydrostatic models.  The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
\begin{align}
\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla.
\end{align}  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
\begin{align} \label{eq:TracerTransport}
\diff{q_i}{t} = 0.
\end{align}

@@

In coordinate-invariant form the prognostic velocity equations may be written in either the Lagrangian or Eulerian frame as
\begin{align} \label{eq:PrognosticVelocity1}
\diff{\vb{u}}{t} = \pdiff{\vb{u}}{t} + \vb{u} \cdot \nabla \vb{u} = - \frac{1}{\rho} \nabla p - 2 \vg{\Omega} \times \vb{u} - \nabla \Phi,
\end{align} where $\vg{\Omega}$ denotes the planetary vorticity vector and $\Phi$ is the geopotential function.  The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.  In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since in an arbitrary coordinate frame this term will give rise to Christoffel symbols associated with derivatives of the vector basis {\color{red}[citation needed]}.  Note that it is common to rewrite the pressure gradient force using the relationship
\begin{align}
- \frac{1}{\rho} \nabla p = - \frac{R_d c_p^\ast}{R^\ast} \theta_v \nabla \pi,
\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).  A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.  This gives rise to the vector-invariant form,
\begin{align} \label{eq:PrognosticVelocity2}
\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}. 
\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis that would emerge in (\ref{eq:PrognosticVelocity1}).  In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
\begin{align}
\pdiff{}{t} (\rho \vb{u}) = - \nabla \cdot (\vb{u} \otimes \vb{u} + \mathcal{I} p) - 2 \vg{\Omega} \times (\rho \vb{u}) - \rho \nabla \Phi,
\end{align} where $\vb{u} \otimes \vb{u}$ denotes the outer product and $\mathcal{I}$ is the identity matrix.

@@

It is also common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.  Decomposing the momentum transport term into horizontal and vertical components gives
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}.
\end{align}  The first term in this expression admits the relationships
\begin{align}
\left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \\
(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right]
\end{align} where $\zeta_h = (\nabla \times \vb{u}) \cdot \vb{k} = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.  Note that this equation does incorporate metric terms associated with horizontal advection of $\vb{k}$ which must be accounted for.

{\color{blue}
\noindent \textbf{Derivation.}
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + \vb{u}_h \cdot \nabla (w \vb{k}) + w \vb{k} \cdot \nabla \vb{u}_h + w \vb{k} \cdot \nabla (w \vb{k}) \\
=& \vb{u}_h \cdot \nabla \vb{u}_h + \nabla( \vb{u}_h \cdot (w \vb{k}) ) - (w \vb{k}) \times (\nabla \times \vb{u}_h) - \vb{u}_h \times (\nabla \times w \vb{k}) + \nabla ( \tfrac{1}{2} w^2) + ((\nabla \times (w \vb{k})) \times (w \vb{k})) \\
=& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}
\end{align} where we have used
\begin{align}
(\nabla \times (w \vb{k})) \times (w \vb{k}) =& (w (\nabla \times \vb{k}) + (\nabla w) \times \vb{k}) \times (w \vb{k}) = - w \vb{k} \times (\nabla w \times \vb{k}) = (- w \vb{k} \cdot \vb{k}) \nabla w - (- w \vb{k} \cdot \nabla w) \vb{k} \\
- \vb{u}_h \times (\nabla \times w \vb{k}) =& - \vb{u}_h \times (w (\nabla \times \vb{k}) + (\nabla w) \times \vb{k}) = - \vb{u}_h \times (\nabla w \times \vb{k}) = (\vb{u}_h \cdot \nabla w) \vb{k}
\end{align}}

Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1}) $\cdot \vb{k}$, leading to
\begin{align} \label{eq:PrognosticWEquationK2}
\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u} \cdot \nabla w - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{\Phi}{z} - (2 \vg{\Omega} \times \vb{u}) \cdot \vb{k}
\end{align}  and
\begin{align} \label{eq:PrognosticUhEquationK2}
\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - (\nabla \times \vb{u}_h) \times (w \vb{k}) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z.
\end{align}  Note that under the shallow atmosphere approximation, the metric term $\vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})$ in (\ref{eq:PrognosticWEquationK2}) is set equal to zero in accordance with \cite{phillips1966equations}.

@@

The dynamical equations are now formulated in terms of the vertical coordinate $s(t,\vb{x},z)$ with $\partial s / \partial z \neq 0$ everywhere, i.e. following \cite{kasahara1974various} (hereafter K74).  Since $\vb{x}$ and $t$ are shared between the two coordinate systems, the chain rule can be applied to obtain expressions
\begin{align} \label{eq:ChainRule}
\pdiff{}{z} =& \pdiff{s}{z} \pdiff{}{s}, & \nabla_s =& \nabla_z + (\nabla_s z) (\vb{k} \cdot \nabla), & \left( \pdiff{}{t} \right)_s =& \pdiff{}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla),
\end{align} which correspond to derivatives in the vertical, in the horizontal and in time.  This final expression is used to describe the rate of change of a quantity on $s$ surfaces.  When applied to $s$, these operators yields the useful identities
\begin{align} \label{eq:NablaZSIdentity}
\nabla_z s = - \left( \pdiff{s}{z} \right) \nabla_s z, \qquad \mbox{and} \qquad \pdiff{s}{t} = - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s.
\end{align}  From here (\ref{eq:ChainRule}) also gives rise to
\begin{align} \label{eq:HorizontalChainRule}
\nabla_z = \nabla_s - \pdiff{s}{z} (\nabla_s z) \pdiff{}{s}.
\end{align}  This equation can be used directly to rewrite (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2}) in terms of derivatives over $s$.  Note that the operators $\nabla_z$ and $\nabla_s$ are usually introduced in the context of 2D flows, however the construction described here has the advantage of working seamlessly in a 3D context, while admitting the properties $\vb{k} \cdot \nabla_z = 0$ and $\vb{k} \cdot \nabla_s = 0$.

The generalized velocity $\dot{s}$ following a fluid parcel is defined by
\begin{align} \label{eq:DefinitionDotS}
\dot{s} \equiv \diff{s}{t} = \pdiff{s}{t} + \vb{u} \cdot \nabla s = \vb{u}_h \cdot \nabla_z s + \left[ w - \left( \pdiff{z}{t} \right)_s \right] \pdiff{s}{z}.
\end{align}  Then using (\ref{eq:ChainRule}) and (\ref{eq:DefinitionDotS}) to rewrite (\ref{eq:MaterialDerivativeScalar}) gives an expression for the material derivative for scalars on $s$ surfaces,
\begin{align}
\diff{A}{t} =& \left( \pdiff{A}{t} \right)_s - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s \pdiff{A}{s} + \vb{u}_h \cdot \left[ \nabla_s A + (\nabla_z s) \pdiff{A}{s} \right] + w \pdiff{s}{z} \pdiff{A}{s} \\
=& \left( \pdiff{A}{t} \right)_s  + \vb{u}_h \cdot \nabla_s A + \dot{s} \pdiff{A}{s}  \label{eq:MaterialDerivativeS}
\end{align}

The prognostic equation for vertical velocity from (\ref{eq:VerticalVelocityOrthogonal}) is then reformulated using (\ref{eq:ChainRule}),
\begin{align} \label{eq:VerticalVelocityOrthogonalS}
\left(\pdiff{w}{t} \right)_s =& \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K + \Phi) \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}_h \right] \cdot \vb{k}.
\end{align}  Similarly, the prognostic equation for horizontal velocity from (\ref{eq:HorizontalVelocityOrthogonal}) is reformulated as
\begin{align} \label{eq:HorizontalVelocityOrthogonalS}
\left( \pdiff{\vb{u}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{u}_h).
\end{align}  Note that the last term of this expression can then be expanded using
\begin{align}
(\vb{k} \cdot \nabla \vb{u}_h) =& - \vb{u}_h \cdot \nabla \vb{k} - \vb{k} \times (\nabla \times \vb{u}_h),
\end{align} where under the shallow-atmosphere approximation $(\vb{u}_h \cdot \nabla \vb{k})$ is typically dropped.

The prognostic equations utilizing horizontal kinetic energy $K_2$ in place of $K$ are similarly derived from (\ref{eq:PrognosticWEquationK2}),
\begin{align} \label{eq:PrognosticWEquationK2S}
\left( \pdiff{w}{t} \right)_s =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u}_h \cdot \nabla w + \left( \pdiff{s}{z} \right) \left\{ \left[ \left( \pdiff{z}{t} \right)_s - w \right] \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) \right\} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k},
\end{align} and from (\ref{eq:PrognosticUhEquationK2}),
\begin{align} \label{eq:PrognosticUhEquationK2S}
\left( \pdiff{\vb{u}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}).
\end{align}  Observe that both of these equations are simplified if $(\partial z / \partial t)_s = w$.

%{\color{red}[Is there an easy way to write down the curl (specifically relative vorticity) in arbitrary vertical coordinates?]}

\subsection{Covariant Formulation}

>>>>

%\Author[18]{Elijah}{Goodfriend}
%\Author[18]{Hans}{Johansen}

@@

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical. DYNAMICO's design uniquely combines:
 \begin{itemize}
\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry,\citep{Dubos2015DYNAMICO10}
\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the litterature,\citep{thuburn2009numerical,ringler2010unified}, and yielding, to a certain extent, an energy-conserving spatial discretization, and
\item a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation). 
\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

The GEM model, which is used for operational forecasting at Environment Canada, uses the Yin-Yang grid and horizontal discretization on an Arakawa C grid.  The vertical coordinate is of a log-hydrostatic-pressure type and vertical discretization is based on the Charney-Phillips grid.  A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.  A scalar viscosity is employed for wind components and tracers 
via an application of a scalar Laplacian.  Viscosity operations are applied after the completion of the dynamic time step. {\color{red}[Perhaps expand this text a little bit?]}

@@

The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} uses an Arakawa C-grid built on a centroidal Voronoi tessellation, in conjunction with the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.  Advection terms are nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.  In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.  The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.  Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.  Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, and fourth-order hyperdiffusion.

@@

The Tempest model \citep{ullrich2014global, guerra2016high} uses a horizontal spectral element discretization and vertical nodal finite volume method with Lorenz staggering based on the cubed-sphere grid with terrain-following height-based coordinate.  The standard Eulerian equations are employed with moist density $\rho$, thermodynamic closure $\theta_v$ and tracer density $\rho q$.  These continuous equations are given in section \ref{sec:TempestEquations}.  The implementation includes both fully explicit time integration, using the UKG53 scheme described in section \ref{sec:UKG53Scheme}, and implicit-explicit options, where horizontal terms are explicitly discretized and vertical terms are treated implicitly.

@@

\section{Moist Nonhydrostatic Equation Sets} \label{sec:EquationSets}

In this section we describe the fluid equations utilized by nonhydrostatic models.  The material derivative is used for  quantities in the Lagrangian frame (following individual air parcels), and is given by
\begin{align}
\diff{}{t} = \pdiff{}{t} + \vb{u} \cdot \nabla.
\end{align}  Note that tracer variables $q_i$, including specific humidity $q$, satisfy the simple Lagrangian relationship
\begin{align} \label{eq:TracerTransport}
\diff{q_i}{t} = 0.
\end{align}

@@

In coordinate-invariant form the prognostic velocity equations may be written in either the Lagrangian or Eulerian frame as
\begin{align} \label{eq:PrognosticVelocity1}
\diff{\vb{u}}{t} = \pdiff{\vb{u}}{t} + \vb{u} \cdot \nabla \vb{u} = - \frac{1}{\rho} \nabla p - 2 \vg{\Omega} \times \vb{u} - \nabla \Phi,
\end{align} where $\vg{\Omega}$ denotes the planetary vorticity vector and $\Phi$ is the geopotential function.  The three terms on the right-hand-side of this expression correspond to pressure gradient, Coriolis, and gravitational force, respectively.  In Eulerian form one must be careful with the treatment of the momentum advection term $\vb{u} \cdot \nabla \vb{u}$, since in an arbitrary coordinate frame this term will give rise to Christoffel symbols associated with derivatives of the vector basis.  Note that it is common to rewrite the pressure gradient force using the relationship
\begin{align}
- \frac{1}{\rho} \nabla p = - \frac{R_d c_p^\ast}{R^\ast} \theta_v \nabla \pi,
\end{align} which follows from (\ref{eq:DiagnosticRelationships1}).  A second form of (\ref{eq:PrognosticVelocity1}) emerges on substituting the vector calculus identity
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \nabla K + \vg{\zeta} \times \vb{u},
\end{align} where $K = \frac{1}{2} (\vb{u} \cdot \vb{u})$ is the 3D Kinetic energy and $\vg{\zeta} = \nabla \times \vb{u}$ is the 3D relative vorticity vector.  This gives rise to the 3D vector-invariant form,
\begin{align} \label{eq:PrognosticVelocity2}
\pdiff{\vb{u}}{t} = - \frac{1}{\rho} \nabla p - \nabla (K + \Phi) - (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}. 
\end{align}  Because no gradients of vectors appear in this equation, it avoids derivatives of the coordinate basis that would emerge in (\ref{eq:PrognosticVelocity1}).  In conjunction with (\ref{eq:ContinuityEquation}), both (\ref{eq:PrognosticVelocity1}) also give rise to the flux-form momentum equations,
\begin{align}
\pdiff{}{t} (\rho \vb{u}) = - \nabla \cdot (\vb{u} \otimes \vb{u} + \mathcal{I} p) - 2 \vg{\Omega} \times (\rho \vb{u}) - \rho \nabla \Phi,
\end{align} where $\vb{u} \otimes \vb{u}$ denotes the outer product and $\mathcal{I}$ is the identity matrix.

@@

Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.  Decomposing the momentum transport term into horizontal and vertical components gives
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}.
\end{align}  The first term in this expression admits the relationships
\begin{align}
\left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \label{eq:UhDotNablaUhRelationship1} \\
(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right] \label{eq:UhDotNablaUhRelationship2}
\end{align} where $\zeta_h = (\nabla \times \vb{u}) \cdot \vb{k} = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.  Note that this equation does incorporate metric terms associated with horizontal advection of $\vb{k}$ which must be accounted for.

Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1})$\cdot \vb{k}$, is
\begin{align} \label{eq:PrognosticWEquationK2}
\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - w \pdiff{w}{z} - \vb{u}_h \cdot \nabla_z w - \pdiff{\Phi}{z} - \frac{1}{\rho} \pdiff{p}{z} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}.
\end{align}  Then subtracting (\ref{eq:PrognosticWEquationK2})$\vb{k}$ from (\ref{eq:PrognosticVelocity1}) gives
\begin{align} \label{eq:PrognosticUhEquationK2}
\pdiff{\vb{u}_h}{t} =& - w (\vb{u}_h \cdot \nabla \vb{k}) - w \pdiff{\vb{u}_h}{z} - \nabla_z (K_2 + \Phi) - \frac{1}{\rho} \nabla_z p - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z.
\end{align}  Note that under the shallow atmosphere approximation, the metric term $\vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})$ in (\ref{eq:PrognosticWEquationK2}) is set equal to zero in accordance with \cite{phillips1966equations}.

@@

The dynamical equations are now formulated in terms of the vertical coordinate $s(t,\vb{x},z)$ with $\partial s / \partial z \neq 0$ everywhere, i.e. following \cite{kasahara1974various} (hereafter K74).  Since $\vb{x}$ and $t$ are shared between the two coordinate systems, the chain rule can be applied to obtain expressions
\begin{align} \label{eq:ChainRule}
\pdiff{}{z} =& \pdiff{s}{z} \pdiff{}{s}, & \nabla_s =& \nabla_z + (\nabla_s z) (\vb{k} \cdot \nabla), & \left( \pdiff{}{t} \right)_s =& \pdiff{}{t} + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla),
\end{align} which correspond to derivatives in the vertical, in the horizontal and in time.  This final expression is used to describe the rate of change of a quantity on $s$ surfaces.  These operators then yield the useful identities
\begin{align} \label{eq:NablaZSIdentity}
\pdiff{s}{z} =& \left( \pdiff{z}{s} \right)^{-1}, & \nabla_z s =& - \left( \pdiff{s}{z} \right) \nabla_s z, & \pdiff{s}{t} =& - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s.
\end{align}  From here (\ref{eq:ChainRule}) also gives rise to
\begin{align} \label{eq:HorizontalChainRule}
\nabla_z = \nabla_s - \pdiff{s}{z} (\nabla_s z) \pdiff{}{s},
\end{align} which can be used directly to rewrite (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2}) in terms of derivatives over $s$.  Note that the operators $\nabla_z$ and $\nabla_s$ are usually introduced in the context of 2D flows, however the construction described here has the advantage of working seamlessly in a 3D context, while admitting the properties $\vb{k} \cdot \nabla_z A = 0$ and $\vb{k} \cdot \nabla_s A = 0$ for any scalar field $A$.

From (\ref{eq:ChainRule}), it can be shown that the 2D divergence on $s$ surfaces (given by K74 eq. (3.17)) is
\begin{align} \label{eq:DivergenceS}
\nabla_s \cdot \vb{u}_h = \nabla_z \cdot \vb{u}_h + \left( \pdiff{s}{z} \right) (\nabla_s z) \cdot \left( \pdiff{\vb{u}_h}{s} \right),
\end{align} and that the 2D curl is given by
\begin{align} \label{eq:CurlS}
\nabla_s \times \vb{u}_h = \nabla_z \times \vb{u}_h + \left( \pdiff{s}{z} \right) (\nabla_s z) \times \left( \pdiff{\vb{u}_h}{s} \right),
\end{align} where $\nabla_z \times \vb{u}_h = \vb{k} (\vb{k} \cdot (\nabla \times \vb{u}_h))$.  Notably, these expressions are valid for both shallow- and deep-atmosphere formulations.

The generalized velocity $\dot{s}$ following a fluid parcel is defined by
\begin{align} \label{eq:DefinitionDotS}
\dot{s} \equiv \diff{s}{t} = \pdiff{s}{t} + \vb{u} \cdot \nabla s = \vb{u}_h \cdot \nabla_z s + \left[ w - \left( \pdiff{z}{t} \right)_s \right] \pdiff{s}{z}.
\end{align}  Then using (\ref{eq:ChainRule}) and (\ref{eq:DefinitionDotS}) to rewrite (\ref{eq:MaterialDerivativeScalar}) gives an expression for the material derivative for scalars on $s$ surfaces,
\begin{align}
\diff{A}{t} =& \left( \pdiff{A}{t} \right)_s - \pdiff{s}{z} \left( \pdiff{z}{t} \right)_s \pdiff{A}{s} + \vb{u}_h \cdot \left[ \nabla_s A + (\nabla_z s) \pdiff{A}{s} \right] + w \pdiff{s}{z} \pdiff{A}{s} \\
=& \left( \pdiff{A}{t} \right)_s  + \vb{u}_h \cdot \nabla_s A + \dot{s} \pdiff{A}{s}  \label{eq:MaterialDerivativeS}
\end{align}  A similar expression arises for vectors, although in this case $\vb{u}_h \cdot \nabla \vb{a} \neq \vb{u}_h \cdot \nabla_z \vb{a}$ implies we cannot use the operator $\nabla_s$ in the form (\ref{eq:ChainRule}), and instead obtain
\begin{align} \label{eq:MaterialDerivativeVectorS}
\diff{\vb{a}}{t} =& \left( \pdiff{\vb{a}}{t} \right)_s + \left[ \vb{u}_h \cdot \nabla \vb{a} + (\vb{u}_h \cdot \nabla_s z) (\vb{k} \cdot \nabla \vb{a}) \right] + \dot{s} \pdiff{\vb{a}}{s}.
\end{align}

%{\color{red}[Is this needed?]} Note that
%\begin{align}
%\nabla A = \nabla_z A + \vb{k} \pdiff{A}{z} = \nabla_s A - \pdiff{s}{z} (\nabla_s z) \pdiff{A}{s} + \vb{k} \pdiff{s}{z} \pdiff{A}{s} = \nabla_s A + ( \vb{k} - \nabla_s z) \pdiff{s}{z} \pdiff{A}{s}
%\end{align}

\subsection{2D Vector Invariant Form}

The prognostic equations utilizing horizontal kinetic energy $K_2$ in place of $K$ are similarly derived by applying (\ref{eq:ChainRule}) to (\ref{eq:PrognosticWEquationK2}), yielding
\begin{align} \label{eq:PrognosticWEquationK2Sa}
\left( \pdiff{w}{t} \right)_s =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u}_h \cdot \nabla w + \left( \pdiff{s}{z} \right) \left\{ \left[ \left( \pdiff{z}{t} \right)_s - w \right] \pdiff{w}{s} - \pdiff{\Phi}{s} - \frac{1}{\rho} \pdiff{p}{s} \right\} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}.
\end{align}  Similarly, from (\ref{eq:PrognosticUhEquationK2}),
\begin{align} \label{eq:PrognosticUhEquationK2Sa}
\left( \pdiff{\vb{u}_h}{t} \right)_s =& - w (\vb{u}_h \cdot \nabla \vb{k}) + \left[ \left( \pdiff{z}{t} \right)_s - w \right] \left( \pdiff{s}{z} \right) \pdiff{\vb{u}_h}{s} - \zeta_h \vb{k} \times \vb{u}_h - \nabla_z (K_2 + \Phi) - \frac{1}{\rho} \nabla_z p - [2 \vg{\Omega} \times \vb{u}]_z.
\end{align}  Observe that both of these equations simplify when $w = (\partial z / \partial t)_s$, \textit{i.e.} model levels are advected with the vertical wind.

An alternative form of these equation can similarly be obtained in terms of $\dot{s}$.  Substituting (\ref{eq:DefinitionDotS}) into (\ref{eq:PrognosticWEquationK2Sa}) then gives
\begin{align} \label{eq:PrognosticWEquationK2Sb}
\left( \pdiff{w}{t} \right)_s =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u}_h \cdot \nabla_s w - \dot{s} \pdiff{w}{s} + \left( \pdiff{s}{z} \right) \left[ - \pdiff{\Phi}{s} - \frac{1}{\rho} \pdiff{p}{s} \right] - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}.
\end{align}  Similarly, substituting (\ref{eq:DefinitionDotS}) into (\ref{eq:PrognosticUhEquationK2Sa}) and using the identity
\begin{align}
(\vb{u}_h \cdot \nabla_s z) \left( \pdiff{s}{z} \right) \pdiff{\vb{u}_h}{s} = (\nabla_s \times \vb{u}_h) \times \vb{u}_h - \zeta_h \vb{k} \times \vb{u}_h + (\nabla_s z) \pdiff{K_2}{z}
\end{align}
 then gives
\begin{align} \label{eq:PrognosticUhEquationK2Sb}
\left( \pdiff{\vb{u}_h}{t} \right)_s =& - w (\vb{u}_h \cdot \nabla \vb{k}) - \nabla_s K_2 - \zeta_s \vb{k} \times \vb{u}_h - \dot{s} \pdiff{\vb{u}_h}{s} - \nabla_z \Phi - \frac{1}{\rho} \nabla_z p - [2 \vg{\Omega} \times \vb{u}]_z,
\end{align} where
\begin{align}
\nabla_s \times \vb{u}_h = \vb{k} \zeta_s, \quad \mbox{and} \quad \zeta_s = \vb{k} \cdot (\nabla_s \times \vb{u}_h).
\end{align}  In this case the vertical advection terms are removed when $\dot{s} = 0$, \textit{i.e.} the vertical coordinate is advected with the 3D wind $\vb{u}$.

Note that under the shallow-atmosphere approximation, the first metric terms (those that include $(\vb{u}_h \cdot \nabla \vb{k})$) in (\ref{eq:PrognosticWEquationK2Sa})-(\ref{eq:PrognosticUhEquationK2Sb}) are typically dropped.

\subsection{3D Vector Invariant Form}
